---
format: live-html
webr:
  packages:
    - dplyr
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| autorun: true
#| echo: false
library(gradethis)
```

:::{.center style="font-size:2em; margin-left:7em;"}
SDS 320E R Demo Lab 
:::


---

[1. Complete the code by filling in the blank such that `1 + _ = 2`]{style="font-size:1.2em;"}

::::{.panel-tabset}

## Exercise
```{webr}
#| caption: Exercise 1
#| exercise: ex_1
1 + _
```


## Hint
::: {.hint exercise="ex_1"}
Remember, at its basics R works just like a calculator.
:::

::::

::: {.solution exercise="ex_1"}
```{webr}
#| exercise: ex_1
#| solution: true
1 + 1
```
:::

```{webr}
#| exercise: ex_1
#| check: true
#| echo: false
gradethis::grade_this_code()
```

---

[2. Fix the following code to make it run.]{style="font-size:1.2em;"}

::::{.panel-tabset}

## Exercise
```{webr}
#| caption: Exercise 2
#| exercise: ex_2
50 + "67"
```


## Hint
::: {.hint exercise="ex_2"}
Is "67" a number, or a string of characters?
:::

::::

::: {.solution exercise="ex_2"}
```{webr}
#| exercise: ex_2
#| solution: true
50 + 67
```
:::

```{webr}
#| exercise: ex_2
#| check: true
#| echo: false
gradethis::grade_this_code()
```


---

[3. Fix the following code to ensure we are using proper variable names.  ]{style="font-size:1.2em;"}

::::{.panel-tabset}

## Exercise
```{webr}
#| caption: a).
#| exercise: ex_3a
1variable <- c(TRUE, FALSE, TRUE, TRUE)
```


## Hint
::: {.hint exercise="ex_3a"}
Where can #'s go when creating a variable name?
:::

::::

::: {.solution exercise="ex_3a"}
```{webr}
#| exercise: ex_3a
#| solution: true
variable <- c(TRUE, FALSE, TRUE, TRUE)
```
:::

```{webr}
#| exercise: ex_3a
#| check: true
#| echo: false
gradethis::grade_this_code()
```

::::{.panel-tabset}

## Exercise
```{webr}
#| caption: b).
#| exercise: ex_3b
_my2ndvariable <- c("cat", "dog", "dog", "cat")
```


## Hint
::: {.hint exercise="ex_3b"}
Where can `_`'s go when creating a variable name?
:::

::::

::: {.solution exercise="ex_3b"}
```{webr}
#| exercise: ex_3b
#| solution: true
my2ndvariable <- c("cat", "dog", "dog", "cat")
```
:::

```{webr}
#| exercise: ex_3b
#| check: true
#| echo: false
gradethis::grade_this_code()
```

---

[4. Consider the variables `A`, `B`, `C`, and `D` below. ]{style="font-size:1.2em;"}

```{r}
A <- 56
B <- 23
C <- 5
D <- 103
```


[Fix the following code that performs mathematical operations on the variables.]{style="font-size:1.2em;"}

::::{.panel-tabset}

## Exercise
```{webr}
#| setup: true
#| exercise: ex_4
A <- 56
B <- 23
C <- 5
D <- 103
```

```{webr}
#| caption: Exercise 4
#| exercise: ex_4
a*B + c + D/4
```


## Hint
::: {.hint exercise="ex_4"}
Does capitalization matter for variable names?
:::

::::

::: {.solution exercise="ex_4"}
```{webr}
#| exercise: ex_4
#| solution: true
A*B + C + D/4
```
:::

```{webr}
#| exercise: ex_4
#| check: true
#| echo: false
gradethis::grade_this_code(
  incorrect = "Not quite — look at the variable names carefully!"
)
```

---

[5. Consider the vector of numbers below.]{style="font-size:1.2em;"}

```{r, eval = FALSE}
c(-1, 1, -2, -1.67, -1.33, -1, -0.67, -0.33, 0, 0.33, 0.67, 1, 1.33, 1.67, 2)
```


[Create a vector named `x_variable` that contains the numbers given above.]{style="font-size:1.2em;"}

::::{.panel-tabset}

## Exercise
```{webr}
#| caption: Exercise 5
#| exercise: ex_5

```


## Hint
::: {.hint exercise="ex_5"}
Creating objects is done by use `<-` or `=`
:::

::::

::: {.solution exercise="ex_5"}
```{webr}
#| exercise: ex_5
#| solution: true
# arrow assignment ----
x_variable <- c(-1, 1, -2, -1.67, -1.33, -1, -0.67, -0.33, 0, 0.33, 0.67, 1, 1.33, 1.67, 2)

# equals assignment ----
x_variable = c(-1, 1, -2, -1.67, -1.33, -1, -0.67, -0.33, 0, 0.33, 0.67, 1, 1.33, 1.67, 2)
```
:::

```{webr}
#| exercise: ex_5
#| check: true
#| echo: false
gradethis::grade_this_code(
)
```

---

[6. I have created another variable called `y_variable`. Using the `plot()` function, plot the `x_variable` from above on the x-axis and the `y_variable` on the y-axis.]{style="font-size:1.2em;"}

::::{.panel-tabset}

## Exercise
```{webr}
#| setup: true
#| exercise: ex_6

x_variable <- c(-1, 1, -2, -1.67, -1.33, -1, -0.67, -0.33, 0, 0.33, 0.67, 1, 1.33, 1.67, 2)
y_variable <- c(1, 1, -0.5, -0.81, -1.06, -1.25, -1.39, -1.47, -1.5, -1.47, -1.39, -1.25, -1.06, -0.81, -0.5)
```

```{webr}
#| caption: Exercise 6
#| exercise: ex_6
plot()
```


## Hint
::: {.hint exercise="ex_6"}
Make sure you entered the variable names correctly and have the x and y variables in the right order. 
:::

::::

::: {.solution exercise="ex_6"}
```{webr}
#| exercise: ex_6
#| solution: true
plot(x_variable, y_variable)
```
:::

```{webr}
#| exercise: ex_6
#| check: true
#| echo: false
gradethis::grade_this_code(
)
```


---

[7. There is a dataset in `R` called starwars. The dataset is given below.]{style="font-size:1.2em;"}

```{r, echo=FALSE, results='hide', message=FALSE}
library(dplyr)
```
```{r}
starwars
```

[You can use the `$` operator to pull out a variable from dataset. For example `my_data$variable3` will pull the variable named `variable3` from `my_data`. ]{style="font-size:1.2em;"}

::::{.panel-tabset}

## Exercise

a). For the `starwars` dataset use the `$` to pull out the hair color variable. 

```{webr}
#| setup: true
#| exercise: ex_7a
library(dplyr)
```

```{webr}
#| caption: 
#| exercise: ex_7a
starwars
```


## Hint
::: {.hint exercise="ex_7a"}
Did you spell the variable correctly?
:::

::::

::: {.solution exercise="ex_7a"}
```{webr}
#| exercise: ex_7a
#| solution: true
starwars$hair_color
```
:::

```{webr}
#| exercise: ex_7a
#| check: true
#| echo: false
gradethis::grade_this_code(
)
```

---

[You can use also use `[row #'s you want, column #'s you want]` to pull specific rows and columns from a dataset. For example, to pull the rows 5 and 6 and columns 10, 11, 12 from `my_data`, we would do `my_data[c(5, 6), c(10, 11, 12)]`. ]{style="font-size:1.2em;"}

::::{.panel-tabset}

## Exercise

b). `homeworld` is the 10th variable, Darth Maul is the 43rd observation. Type in the code to determine Darth Mauls homeworld. 

```{webr}
#| setup: true
#| exercise: ex_7b
library(dplyr)
```

```{webr}
#| caption: 
#| exercise: ex_7b
starwars
```


## Hint
::: {.hint exercise="ex_7b"}
Did you use the right rows and columns?
:::

::::

::: {.solution exercise="ex_7b"}
```{webr}
#| exercise: ex_7b
#| solution: true

# ----
starwars[43, 10]

# ----
starwars[c(43), c(10)]

# ----
starwars[(43), (10)]

# ----
starwars[c(43), 10]

# ----
starwars[43, c(10)]

# ----
starwars[43, (10)]

# ----
starwars[(43), 10]
```
:::

```{webr}
#| exercise: ex_7b
#| check: true
#| echo: false
gradethis::grade_this_code(
)
```
